


\chapter{Epigrams}
The state language is the language of politics and government. The national language is the language of lovemaking.

\separator

I often think English monolingual speakers know nothing of the joy of speaking one’s language, because they’ve never known any other language, and they do not know the full range of capabilities and incapabilities of the English language. They therefore use words most uncarefully. They speak but do not say things. Not because

\separator

A paradigm is a comprehensive interpretation system of facts.

An ideology is a paradigm coupled with commandments.

\separator


Why the Cantonese, with her cuisine so alike with the Japanese in talent and delight, fail to match them in worldly sophistication in the eyes of the peoples of the world?

\separator

We the Cantonese connect with each other by speaking extraordinarily loud, angrily, red facedly, and aggressively, over the dinner table filled to the brim with food


\separator

每隻講出口嘅廣東話時文，都係一粒解放香港嘅子彈。

\separator

我肯定他朝有日粵人一定會有人說：「我們自古以來就是天朝最忠誠地添煩添亂的蠻夷，陳炯名是我們偉大的民族英雄，其身承蒙自趙佗和冼夫人那抗命不屈、誓要獨立的大粵國靈，本來勢必建國，完成大粵人民二千年來的刺秦偉業。

\separator

我哋幾時先至可以用廣東話嚟flirt？我哋幾時先至可以有人用廣東話講到唔肉麻嘅「我愛你」？幾時先至可以好似法國佬、意大利佬、西班牙佬咁用自己嘅話嚟征服天下間嘅女人？我哋幾時先至可以有隻可以噒到乸嘅脷？呼風喚雨，狂花亂墜，講到係乸嘅都心淯淯？廣東話一日未畀到係識講廣東話嘅人到做到花靚，就一日係一隻連人類基本情緒到𠄡嘅話。一隻冇浪漫嘅話，就係一隻冇戀愛，淨係得交配嘅賤話。

\separator 



A language has rights as long as its people think it matters and find it worthwhile to preserve. If its people do not think so, then it all comes down to cultural utilities. If a language is almost dead, or already dead, then it only comes down to cultural utilities. In other words, the more successfully eradicated a language, the more it does not matter. It sounds paradoxical, but the alternative is that all languages matter equally and independently, which is something that appeals to people’s sense of equality but it’s just that and nothing else. In fact it is deeply counterintuitive because it means the rights of the living are to be weighed against the rights of the dead. Do you not see? “Rights of the dead”. How could the dead have any rights? The dead do not have any rights except ceremonialiously through the living. We do not mutiliate the dead because we the dead have the rights not to be mutiliated, but because in life their living connections have the right to not be offended. The dead are merely lumps of matter. They themsleves do not have rights.

\separator

我地必須日本化，做到日本已經成功做咗嘅之餘，我地仲要超越佢哋。我地要成功將我地嘅文化發展到班鬼佬同大陸人仆崩鼻過黎崇拜我地同加入我地，要佢哋為自己可以講到粵語而自豪，要令佢哋自願同興高采烈咁摒棄自己嘅語言，就好似班入日嘅󱛞\footnote{Gaijin}淨係講日文咁。我地仲要令到佢哋自己嘅圈子都內部講粵語，就好似好多honkies abc 同喺英國讀書得耐嘅香港人自己圍位都會講英文。

